To understand why spam classification is important to our lives, we must first understand the spam emails and its impact on daily life and businesses.

\subsection{Different Types of Spam Emails}
\label{subsec:different-types-of-spam-emails}

There are various forms of spam, sent with different intentions and purposes.
But they are commonly grouped into:

\begin{itemize}
    \item Phishing Emails
    \item Email Spoofing
    \item Technical support scams
    \item Current event scams
    \item Marketing/advertising email
    \item Malware scam
\end{itemize}

\subsection{Problems with Spam Emails}
\label{subsec:problems-with-spam-emails}

According to statistics report in 2023, 160 billion spam emails are sent every day, which is 46\% of the 347 billion emails sent daily.
Out of which, the most common type being marketing/advertising emails, which take up around 36\%, followed up with promotional of adult content, around 31.7\% of total spam emails.
Despite scam and fraudulent emails being the least common type, over 70\% of them are phishing emails, which is still over 6 billion phishing emails are being sent to users daily.

A single spam email carbon emission is almost 0.03g of Carbon Dioxide Equivalent (CO$_{2}$e).
With the amount of spam being sent daily, it can easily get nearly 5 tonnes of CO$_{2}$e being released every day.
Additionally, two-thirds of spam receivers have been reported to have their mental health affected due to the number of spam or phishing scams.

For businesses, this spam can be sent as a way to get businesses to invest in nonexistent organizations under the guise of an investment and promise a high payback.
For individuals, it would be under the form of bitcoin investment or for a charitable cause.
Once the money is received, the sender would delete all traces and block the recipient contact.